The curious world of constructive mathematics

### 2022-09-08 algar-lecture-notes.tex
Beginning of list of globalization examples
Number exercises in a more sensible way
Introduction to topos-theoretic phantoms

@@

\newtheorem{exercise}[defn]{Exercise}
\renewcommand{\theenumi}{\alph{enumi}}

@@

@@

>>>>

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

@@

\newtheorem{exercise}{Exercise}[chapter]
\renewcommand{\theenumi}{\alph{enumi}}

@@

\newpage

\begin{tabular}{@{}lp{0.83\textwidth}ll@{}}
  \toprule
  & statement & status \\\midrule
  1a & Every continuous function~$f : [-1,1] \to \RR$ with~$f(-1) < 0$
  and~$f(1) > 0$ has a zero. & \cmark/\xmark \\
  1b & For every continuous function~$f : X \times [-1,1] \to \RR$
  with~$f(\cdot,-1) < 0$ and~$f(\cdot,1) > 0$ there is an open covering~$X =
  \bigcup_i U_i$ such that for each index~$i$, there is a continuous
  zero-picking function~$z : U_i \to [-1,1]$
\end{tabular}

@@

The convergence radius of the Taylor series expansion of~$1/(1-x)$ around the
origin, that is the geometric series
$1 + x + x^2 + \cdots$,
is~$1$. That the convergence radius isn't larger doesn't come too surprising in
view of the fact that the function~$1/(1-x)$ has a non-removable discontinuity
at~$x = 1$. But why is also the convergence radius of the expansion
of~$1/(1+x^2)$,
\[ 1/(1+x^2) = 1 - x^2 + x^4 - x^6 + \cdots, \]
just~$1$? The function~$1/(1+x^2)$ is
perfectly well-defined at~$\pm 1$, infinitely derivable on all of~$\RR$, neatly
decaying for~$x \to \pm\infty$ and doesn't have any singularities.

A rewarding explanation of this phenomenon is nowadays relayed in any course on
complex analysis: The function~$1/(1+x^2)$ does have a singularity at
distance~$1$ from the origin, namely at~$\pm i$ in the complex plane, and the disk
of convergence is only so big that it merely touches the nearest singularity.

This episode is one example of many how the ``imaginary unit''---once
an elusive \emph{mathematical phantom}---affects the story of the real numbers even
though it itself is not one of them. Like many mathematical phantoms before it,
the imaginary unit obtruded its effects so convincingly that eventually we
embraced it and broadened our notion of existence.

Toposes allow us to bring further entities into being, entities which do not
exist in the standard topos but which nevertheless affect the story of the
standard topos. We will focus on:
\begin{enumerate}
\item The \emph{generic ring} -- which cryptically is also a field
\item The \emph{generic prime filter}~$\ppp_\star$ of a given ring~$A$ -- which
XXX
\item The \emph{generic surjection}~$\NN \to X$ -- which exists also when~$X$
is uncountable
\end{enumerate}
The complex numbers can be compiled down to pairs of real numbers and hence
exhibit already in their very construction an intimate connection with the real
numbers; in a similar vein, these topos-theoretic generic models can be
compiled away in that proofs employing them can be unrolled, in a mechanical
fashion even, so as to not mention them.
